Fitting metabolic models to dissolved oxygen data: the Estuarine BAyesian Single-station Estimation method

### 2023-05-09 manu-draft.qmd
adding citations to intro

Methods for estimating metabolic rates in aquatic environments are numerous . Kemp and Testa (2011) and Staehr et al. (2012) summarized the main approaches for quantifying ecosystem metabolism in aquatic systems (including incubations, the open-water method, ecosystem budgets, oxygen isotopes, and inert gases).

Complementing the incubation-based methods are open-water methods (Odum, 1956), which exploit the diel cycle of dissolved oxygen and allow the simultaneous determination of whole-system gross primary production (GPP, equals net primary production + autotrophic respiration), ecosystem respiration (ER, equals heterotrophic + autotrophic respiration), and NEP (equals GPP -- ER). A global synthesis of warm-season, open-water metabolism studies (Hoellein et al., 2013) was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries (Caffrey, 2004); a similar synthesis by Yvon-Durocher et al. (2012) was exclusively based on Caffrey (2004). As such, a unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes. Such a model should be informed by the available data at a given location and have the ability to incorporate prior information on key parameters if available. Bayesian techniques have not been fully leveraged in estuarine applications despite their potential to address the above needs. A potentially useful application developed for streams is the BAyesian Single-station Estimation method.

Application of open-water techniques for estimating metabolism is especially problematic based on the assumptions and poor error constraints of existing methods that rely on continuous measurements of dissolved oxygen. Fundamental to these challenges is the high temporal and spatial variability noted in global syntheses of primary production and respiration. Cloern et al. (2014) found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year. Hopkinson and Smith (2005) focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with wholesystem respiration rates based on the open-water method (Hoellein et al., 2013).

The goal of this paper is to describe the development and application of a new data assimilation method for simultaneously determining gross primary production (P), ecosystem respiration (R), net ecosystem metabolism (NEM), and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, EBASE is compared to metabolic estimates using the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

@@

The development of EBASE represents a new approach for estimating metabolic rates in estuarine settings that leverages existing methods by adapting theoretical components of the underlying mass-balance model using a Bayesian framework. Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty using uninformed prior distributions and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings. A notable limitation of EBASE was an inability to preciseily recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series. Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample with the same metabolic history. Further development of EBASE to accommodate a tidal signal, or exploring alternative methods for detiding dissolved oxygen data, represent promising avenues of future research to more accurately quantify metabolic parameters in estuarines. Regardless, EBASE represents a potentially powerful tool that removes many of the assumptions of existing methods by allowing unknown metabolic parameters to be estimated by the data and can also incorporate existing knowledge for further exploration of estuarine ecosystem properties.

>>>>

Methods for estimating metabolism in aquatic environments are numerous [@kemp2011; @staehr2011]. Chief among these are open-water methods [@Odum56], which exploit the diel cycle of dissolved oxygen and allow the simultaneous determination of whole-system gross primary production (P, equals net primary production + autotrophic respiration), ecosystem respiration (R, equals heterotrophic + autotrophic respiration), and NE (equals P-R). A global synthesis of warm-season, open-water metabolism studies (Hoellein et al., 2013) was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries [@caffrey2004]; a similar synthesis by @yvon-durocher2012 was exclusively based on @caffrey2004. As such, a unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes. Such a model should be informed by the available data at a given location and have the ability to incorporate prior information on key parameters if available. Bayesian techniques have not been fully leveraged in estuarine applications despite their potential to address the above needs. A potentially useful application developed for streams is the BAyesian Single-station Estimation method....

Application of open-water techniques for estimating metabolism is especially problematic based on the assumptions and poor error constraints of existing methods that rely on continuous measurements of dissolved oxygen. Fundamental to these challenges is the high temporal and spatial variability noted in global syntheses of primary production and respiration. Cloern et al. (2014) found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year. Hopkinson and Smith (2005) focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with wholesystem respiration rates based on the open-water method (Hoellein et al., 2013). Tidal variation...

The goal of this paper is to describe the development and application of a new data assimilation method for simultaneously determining gross primary production, ecosystem respiration, net ecosystem metabolism (NEM), and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, EBASE is compared to metabolic estimates using the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

@@

The development of EBASE represents a new approach for estimating metabolic rates in estuarine settings that leverages existing methods by adapting theoretical components of the underlying mass-balance model using a Bayesian framework. Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty using uninformed prior distributions and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings. A notable limitation of EBASE was an inability to precisely recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series. Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample having a continuous metabolic history. Further development of EBASE to accommodate a tidal signal, or exploring alternative methods for detiding dissolved oxygen data, are promising avenues of future research to more accurately quantify metabolic parameters in estuaries. Overall, EBASE represents a potentially powerful tool that removes many of the assumptions of existing methods by allowing unknown metabolic parameters to be estimated by the data and can also incorporate existing knowledge for further exploration of estuarine ecosystem properties.
